The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America

### 2024-01-17 manuscript-1-old/manuscript-latin-countries.tex
compiling-manuscript

% configure pdf metadata and link handling
\hypersetup{
	pdfauthor={Carmen Cabrera-Arnau},
	pdftitle={The mid-term impact of COVID-19 on human mobility patterns in Latin American countries},
	pdfsubject={Title},
	pdfkeywords={Keywords},
	pdffitwindow=true,
	breaklinks=true,
	colorlinks=false,
	pdfborder={0 0 0}}

\title{The mid-term impact of COVID-19 on human mobility patterns in Latin American countries\footnote{\textbf{Citation}: Cabrera-Arnau, C., González-Leonardo, M., Nasuto, A., Neville, R., Rowe, F. (2023). The mid-term impact of COVID-19 on human mobility patterns in Latin American countries}}
\author[1]{Carmen Cabrera-Arnau \thanks{\textit{Corresponding author}: c.cabrera-arnau@liverpool.ac.uk}}
\author[2]{Miguel González-Leonardo}
\author[1]{Andrea Nasuto}
\author[1]{Ruth Neville}
\author[1]{Francisco Rowe}
\affil[1]{Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, United Kingdom}
\affil[2]{Center for Demographic, Urban and Environmental Studies, El Colegio de México (COLMEX), Mexico City, Mexico}

@@

Recent empirical studies, predominantly from countries in the Global North, have shown that the COVID-19 pandemic disrupted human mobility at different spatial scales, but particularly in big cities, initially identified as epicentres of infections. Lockdowns, remote work, and online education decreased the demand for commuting and urban living, resulting in an "urban exodus". Despite the existing evidence, little is known about whether counterurbanisation movements have unfolded similarly in the Global South. In this study, we use anonymised, high-resolution location data from Facebook users to answer the following research questions: i) to what extent are people leaving cities? ii) are there variations across countries? iii) are there changes in mobility patterns over time? iv) how long do these changes last? The data is aggregated into spatial units according to the Bing Maps Tile System and temporally aggregated into 8-hour time windows for a period of approximately 2 years. The methodology involves three steps. Firstly, we classify the Bing tiles into different population density categories according to population data from WorldPop. Secondly, we quantify and analyse the number and characteristics of movements between different locations in each country. Finally, we use statistical modelling and spatial interaction models to assess how the population density category at the origin and destination affects the strength of the flows at different points in time. We observe that the strength of flows between Bing tiles belonging to different population density categories was systematically reduced in the early days of the pandemic and, even though by 2022 it seemed to have increased, it never returned to pre-pandemic levels. In the early days of COVID-19, we also find a systematically increase above pre-pandemic levels of the strength of the flows between Bing tiles belonding to the the same population density category. As of 2022, the frequency of these movements within population density classes have decreased with respect to 2020.

@@

\begin{figure}

{\centering \includegraphics[width=6.51042in,height=6.66667in]{images/Map_density_histogram.png}

@@

\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aromuxed2023}{}}%
Aromí, Daniel, María Paula Bonel, Julian Cristia, Martín Llada, Juan
Pereira, Xiomara Pulido, and Julieth Santamaria. 2023. {``\#StayAtHome:
Social Distancing Policies and Mobility in Latin America and the
Caribbean.''} \emph{Economía} 22 (1).
\url{https://doi.org/10.31389/eco.4}.

\leavevmode\vadjust pre{\hypertarget{ref-BBCnews2023}{}}%
BBC news. 2023. {``Unemployment: Who Are the Millions of Britons Not
Working?''} \url{https://www.bbc.co.uk/news/business-52660591}.

\leavevmode\vadjust pre{\hypertarget{ref-bell2015}{}}%
Bell, Martin, Elin Charles-Edwards, Philipp Ueffing, John Stillwell,
Marek Kupiszewski, and Dorota Kupiszewska. 2015. {``Internal Migration
and Development: Comparing Migration Intensities Around the World.''}
\emph{Population and Development Review} 41 (1): 33--58.
\url{https://doi.org/10.1111/j.1728-4457.2015.00025.x}.

\leavevmode\vadjust pre{\hypertarget{ref-bernard2017}{}}%
Bernard, Aude, Francisco Rowe, Martin Bell, Philipp Ueffing, and Elin
Charles-Edwards. 2017. {``Comparing Internal Migration Across the
Countries of Latin America: A Multidimensional Approach.''} Edited by
Osman Alimamy Sankoh. \emph{PLOS ONE} 12 (3): e0173895.
\url{https://doi.org/10.1371/journal.pone.0173895}.

\leavevmode\vadjust pre{\hypertarget{ref-bhadra2020}{}}%
Bhadra, Arunava, Arindam Mukherjee, and Kabita Sarkar. 2020. {``Impact
of Population Density on Covid-19 Infected and Mortality Rate in
India.''} \emph{Modeling Earth Systems and Environment} 7 (1): 623--29.
\url{https://doi.org/10.1007/s40808-020-00984-7}.

\leavevmode\vadjust pre{\hypertarget{ref-blustein2020}{}}%
Blustein, David L., Ryan Duffy, Joaquim A. Ferreira, Valerie
Cohen-Scali, Rachel Gali Cinamon, and Blake A. Allan. 2020.
{``Unemployment in the Time of COVID-19: A Research Agenda.''}
\emph{Journal of Vocational Behavior} 119 (June): 103436.
\url{https://doi.org/10.1016/j.jvb.2020.103436}.

\leavevmode\vadjust pre{\hypertarget{ref-borsdorf2003}{}}%
Borsdorf, Axel. 2003. {``Cómo Modelar El Desarrollo y La Dinámica de La
Ciudad Latinoamericana.''} \emph{EURE (Santiago)} 29 (86).
\url{https://doi.org/10.4067/s0250-71612003008600002}.

\leavevmode\vadjust pre{\hypertarget{ref-branduxe9n2020}{}}%
Brandén, Maria, Siddartha Aradhya, Martin Kolk, Juho Härkönen, Sven
Drefahl, Bo Malmberg, Mikael Rostila, Agneta Cederström, Gunnar
Andersson, and Eleonora Mussino. 2020. {``Residential Context and
COVID-19 Mortality Among Adults Aged 70 Years and Older in Stockholm: A
Population-Based, Observational Study Using Individual-Level Data.''}
\emph{The Lancet Healthy Longevity} 1 (2): e80--88.
\url{https://doi.org/10.1016/s2666-7568(20)30016-7}.

\leavevmode\vadjust pre{\hypertarget{ref-Brea03}{}}%
Brea, J. 2003. {``Population Dynamics in Latin America.''}
\emph{Population Bulletin} 58: 3--36.
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-chuxe1vezgalindo2016}{}}%
Chávez Galindo, Ana María, Jorge Rodríguez Vignoli, Mario Acuña, Jorge
Barquero, Daniel Macadar, José Marcos Pinto da Cunha, and Jaime Sobrino.
2016. {``Migración Interna y Cambios Metropolitanos.''} \emph{Revista
Latinoamericana de Población} 10 (18): 7--41.
\url{https://doi.org/10.31406/relap2016.v10.i1.n18.1}.

\leavevmode\vadjust pre{\hypertarget{ref-Chen2004}{}}%
Chen, Wenhong, and Barry Wellman. 2004. {``The Global Digital Divide -
Within and Between Countries.''} \emph{IT \& Society} 1: 39--45.
\url{http://www.ifs.tuwien.ac.at/~dieter/teaching/GmA/Chen2004.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-fielding2021}{}}%
Fielding, Tony, and Yoshitaka Ishikawa. 2021. {``COVID{-}19 and
Migration: A Research Note on the Effects of COVID{-}19 on Internal
Migration Rates and Patterns in Japan.''} \emph{Population, Space and
Place} 27 (6). \url{https://doi.org/10.1002/psp.2499}.

\leavevmode\vadjust pre{\hypertarget{ref-firebaugh1979}{}}%
Firebaugh, Glenn. 1979. {``Structural Determinants of Urbanization in
Asia and Latin America, 1950- 1970.''} \emph{American Sociological
Review} 44 (2): 199. \url{https://doi.org/10.2307/2094505}.

\leavevmode\vadjust pre{\hypertarget{ref-florida2021}{}}%
Florida, Richard, Andrés Rodríguez-Pose, and Michael Storper. 2021.
{``Critical Commentary: Cities in a Post-COVID World.''} \emph{Urban
Studies} 60 (8): 1509--31.
\url{https://doi.org/10.1177/00420980211018072}.

\leavevmode\vadjust pre{\hypertarget{ref-ghosh2020}{}}%
Ghosh, Somenath, Pallabi Seth, and Harsha Tiwary. 2020. {``How Does
Covid-19 Aggravate the Multidimensional Vulnerability of Slums in India?
A Commentary.''} \emph{Social Sciences \& Humanities Open} 2 (1):
100068. \url{https://doi.org/10.1016/j.ssaho.2020.100068}.

\leavevmode\vadjust pre{\hypertarget{ref-ginsburg2022}{}}%
Ginsburg, Carren, Mark A. Collinson, F. Xavier Gómez-Olivé, Sadson
Harawa, Chantel F. Pheiffer, and Michael J. White. 2022. {``The Impact
of COVID-19 on a Cohort of Origin Residents and Internal Migrants from
South Africa's Rural Northeast.''} \emph{SSM - Population Health} 17
(March): 101049. \url{https://doi.org/10.1016/j.ssmph.2022.101049}.

\leavevmode\vadjust pre{\hypertarget{ref-Glaeser20}{}}%
Glaeser, Edward L. 2020. {``Cities and Pandemics Have a Long History.
City Journal.''}
\url{https://www.city-journal.org/article/cities-and-pandemics-have-a-long-history}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lezollino2006}{}}%
González Ollino, Daniela, and Jorge Rodríguez Vignoli. 2006. {``Spatial
Redistribution and Internal Migration of the Populationin Chile over the
Past 35 Years (1965-20).''} \emph{Estudios Demográficos y Urbanos} 21
(2): 369. \url{https://doi.org/10.24201/edu.v21i2.1253}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022b}{}}%
González-Leonardo, Miguel, Antonio López-Gay, Niall Newsham, Joaquín
Recaño, and Francisco Rowe. 2022. {``Understanding Patterns of Internal
Migration During the COVID{-}19 Pandemic in Spain.''} \emph{Population,
Space and Place} 28 (6). \url{https://doi.org/10.1002/psp.2578}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2023}{}}%
González-Leonardo, Miguel, Michaela Potančoková, Dilek Yildiz, and
Francisco Rowe. 2023. {``Quantifying the Impact of COVID-19 on
Immigration in Receiving High-Income Countries.''} Edited by Ricky Chee
Jiun Chia. \emph{PLOS ONE} 18 (1): e0280324.
\url{https://doi.org/10.1371/journal.pone.0280324}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022a}{}}%
González-Leonardo, Miguel, Francisco Rowe, and Alberto
Fresolone-Caparrós. 2022. {``Rural Revival? The Rise in Internal
Migration to Rural Areas During the COVID-19 Pandemic. Who Moved and
Where?''} \emph{Journal of Rural Studies} 96 (December): 332--42.
\url{https://doi.org/10.1016/j.jrurstud.2022.11.006}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022}{}}%
González-Leonardo, Miguel, and Jeroen Spijker. 2022. {``The Impact of
Covid-19 on Demographic Components in Spain, 2020{\textendash}31: A
Scenario Approach.''} \emph{Population Studies}, November, 1--17.
\url{https://doi.org/10.1080/00324728.2022.2138521}.

\leavevmode\vadjust pre{\hypertarget{ref-graizbord2007}{}}%
Graizbord, Boris, and Beatriz Acuña. 2007. {``Movilidad Residencial En
La Ciudad de México / Residential Mobility in Mexico City.''}
\emph{Estudios Demográficos y Urbanos} 22 (2): 291.
\url{https://doi.org/10.24201/edu.v22i2.1281}.

\leavevmode\vadjust pre{\hypertarget{ref-green2021}{}}%
Green, Mark, Frances Darlington Pollock, and Francisco Rowe. 2021.
{``New Forms of Data and New Forms of Opportunities to Monitor and
Tackle a Pandemic.''} In, 423--29. Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-70179-6_56}.

\leavevmode\vadjust pre{\hypertarget{ref-haslag2021}{}}%
Haslag, Peter H., and Daniel Weagley. 2021. {``From L.A. To Boise: How
Migration Has Changed During the COVID-19 Pandemic.''} \emph{SSRN
Electronic Journal}. \url{https://doi.org/10.2139/ssrn.3808326}.

\leavevmode\vadjust pre{\hypertarget{ref-huremovic2019}{}}%
Huremović, Damir. 2019. {``Brief History of Pandemics (Pandemics
Throughout History).''} In, 7--35. Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-15346-5_2}.

\leavevmode\vadjust pre{\hypertarget{ref-internat2021}{}}%
{``International Migration 2020: Highlights.''} 2021, January.
\url{https://doi.org/10.18356/9789210052689}.

\leavevmode\vadjust pre{\hypertarget{ref-irudayarajan2020}{}}%
Irudaya Rajan, S., P. Sivakumar, and Aditya Srinivasan. 2020. {``The
COVID-19 Pandemic and Internal Labour Migration in India: A {`}Crisis of
Mobility{'}.''} \emph{The Indian Journal of Labour Economics} 63 (4):
1021--39. \url{https://doi.org/10.1007/s41027-020-00293-8}.

\leavevmode\vadjust pre{\hypertarget{ref-janoschka2002}{}}%
Janoschka, Michael. 2002. {``El Nuevo Modelo de La Ciudad
Latinoamericana: Fragmentación y Privatización.''} \emph{EURE
(Santiago)} 28 (85).
\url{https://doi.org/10.4067/s0250-71612002008500002}.

\leavevmode\vadjust pre{\hypertarget{ref-kotsubo2022}{}}%
Kotsubo, Masaki, and Tomoki Nakaya. 2022. {``Trends in Internal
Migration in Japan, 2012{\textendash}2020: The Impact of the COVID{-}19
Pandemic.''} \emph{Population, Space and Place} 29 (4).
\url{https://doi.org/10.1002/psp.2634}.

\leavevmode\vadjust pre{\hypertarget{ref-lattes95}{}}%
Lattes, A. 1995. {``Urbanización, Crecimiento Urbano y Migraciones En
América Latina. Población y Desarrollo: Tendencias y Desafíos.''}
\emph{Notas de Población} 62: 211--60.
\url{https://repositorio.cepal.org/handle/11362/38594}.

\leavevmode\vadjust pre{\hypertarget{ref-lattes2017}{}}%
Lattes, Alfredo E., Jorge Rodríguez, and Miguel Villa. 2017.
{``Population Dynamics and Urbanization in Latin America: Concepts and
Data Limitations.''} In, 89--111. Routledge.
\url{https://doi.org/10.4324/9781315248073-5}.

\leavevmode\vadjust pre{\hypertarget{ref-lucchini2023}{}}%
Lucchini, Lorenzo, Ollin Langle-Chimal, Lorenzo Candeago, Lucio Melito,
Alex Chunet, Aleister Montfort, Bruno Lepri, Nancy Lozano-Gracia, and
Samuel P. Fraiberger. 2023. {``Socioeconomic Disparities in Mobility
Behavior During the COVID-19 Pandemic in Developing Countries.''}
\url{https://doi.org/10.48550/ARXIV.2305.06888}.

\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\leavevmode\vadjust pre{\hypertarget{ref-McKinsey21}{}}%
McKinsey \& Company. 2021. {``What Executives Are Saying about the
Future of Hybrid Work.''}
\url{https://www.mckinsey.com/capabilities/people-and-organizational-performance/our-insights/what-executives-are-saying-about-the-future-of-hybrid-work}.

\leavevmode\vadjust pre{\hypertarget{ref-ILO2020}{}}%
Nair, Shabari, and Divya Verma. 2020. {``A Policy Framework for India's
Covid-19 Migration. International Labour Organization.''}
\url{https://www.ilo.org/newdelhi/info/public/fs/WCMS_746654/lang--en/index.htm}.

\leavevmode\vadjust pre{\hypertarget{ref-nathan2020}{}}%
Nathan, Max, and Henry Overman. 2020. {``Will Coronavirus Cause a Big
City Exodus?''} \emph{Environment and Planning B: Urban Analytics and
City Science} 47 (9): 1537--42.
\url{https://doi.org/10.1177/2399808320971910}.

\leavevmode\vadjust pre{\hypertarget{ref-UNpopulation19}{}}%
Nations", "United. 2019. {``World Urbanization Prospects. The 2018
Revision.''}
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
Nouvellet, Pierre, Sangeeta Bhatia, Anne Cori, Kylie E. C. Ainslie, Marc
Baguelin, Samir Bhatt, Adhiratha Boonyasiri, et al. 2021. {``Reduction
in Mobility and COVID-19 Transmission.''} \emph{Nature Communications}
12 (1). \url{https://doi.org/10.1038/s41467-021-21358-2}.

\leavevmode\vadjust pre{\hypertarget{ref-oecdreg}{}}%
\emph{OECD Regions and Cities at a Glance}. n.d. OECD.
\url{https://doi.org/10.1787/26173212}.

\leavevmode\vadjust pre{\hypertarget{ref-oecdsoc}{}}%
{``OECD Social, Employment and Migration Working Papers.''} n.d.
\url{https://doi.org/10.1787/1815199x}.

>>>>

% configure pdf metadata and link handling
\hypersetup{
	pdfauthor={Carmen Cabrera-Arnau},
	pdftitle={The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America},
	pdfsubject={Title},
	pdfkeywords={Keywords},
	pdffitwindow=true,
	breaklinks=true,
	colorlinks=false,
	pdfborder={0 0 0}}

\title{The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America}
\author[1]{Carmen Cabrera-Arnau \thanks{\textit{Corresponding author}: c.cabrera-arnau@liverpool.ac.uk}}
\author[2]{Miguel González-Leonardo}
\author[1]{Andrea Nasuto}
\author[1]{Ruth Neville}
\author[1]{Francisco Rowe}
\affil[1]{Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, United Kingdom}
\affil[2]{Center for Demographic, Urban and Environmental Studies, El Colegio de México (COLMEX), Mexico City, Mexico}

@@

@@

\begin{figure}[H]

{\centering \includegraphics[width=6.51042in,height=6.66667in]{figures/Map_density_histogram.png}

@@

\end{figure}%

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-aromuxed2023}
Aromí, Daniel, María Paula Bonel, Julian Cristia, Martín Llada, Juan
Pereira, Xiomara Pulido, and Julieth Santamaria. 2023. {``\#StayAtHome:
Social Distancing Policies and Mobility in Latin America and the
Caribbean.''} \emph{Economía} 22 (1).
\url{https://doi.org/10.31389/eco.4}.

\bibitem[\citeproctext]{ref-BBCnews2023}
BBC news. 2023. {``Unemployment: Who Are the Millions of Britons Not
Working?''} \url{https://www.bbc.co.uk/news/business-52660591}.

\bibitem[\citeproctext]{ref-bell2015}
Bell, Martin, Elin Charles-Edwards, Philipp Ueffing, John Stillwell,
Marek Kupiszewski, and Dorota Kupiszewska. 2015. {``Internal Migration
and Development: Comparing Migration Intensities Around the World.''}
\emph{Population and Development Review} 41 (1): 33--58.
\url{https://doi.org/10.1111/j.1728-4457.2015.00025.x}.

\bibitem[\citeproctext]{ref-bernard2017}
Bernard, Aude, Francisco Rowe, Martin Bell, Philipp Ueffing, and Elin
Charles-Edwards. 2017. {``Comparing Internal Migration Across the
Countries of Latin America: A Multidimensional Approach.''} Edited by
Osman Alimamy Sankoh. \emph{PLOS ONE} 12 (3): e0173895.
\url{https://doi.org/10.1371/journal.pone.0173895}.

\bibitem[\citeproctext]{ref-bhadra2020}
Bhadra, Arunava, Arindam Mukherjee, and Kabita Sarkar. 2020. {``Impact
of Population Density on Covid-19 Infected and Mortality Rate in
India.''} \emph{Modeling Earth Systems and Environment} 7 (1): 623--29.
\url{https://doi.org/10.1007/s40808-020-00984-7}.

\bibitem[\citeproctext]{ref-blustein2020}
Blustein, David L., Ryan Duffy, Joaquim A. Ferreira, Valerie
Cohen-Scali, Rachel Gali Cinamon, and Blake A. Allan. 2020.
{``Unemployment in the Time of COVID-19: A Research Agenda.''}
\emph{Journal of Vocational Behavior} 119 (June): 103436.
\url{https://doi.org/10.1016/j.jvb.2020.103436}.

\bibitem[\citeproctext]{ref-borsdorf2003}
Borsdorf, Axel. 2003. {``Cómo Modelar El Desarrollo y La Dinámica de La
Ciudad Latinoamericana.''} \emph{EURE (Santiago)} 29 (86).
\url{https://doi.org/10.4067/s0250-71612003008600002}.

\bibitem[\citeproctext]{ref-branduxe9n2020}
Brandén, Maria, Siddartha Aradhya, Martin Kolk, Juho Härkönen, Sven
Drefahl, Bo Malmberg, Mikael Rostila, Agneta Cederström, Gunnar
Andersson, and Eleonora Mussino. 2020. {``Residential Context and
COVID-19 Mortality Among Adults Aged 70 Years and Older in Stockholm: A
Population-Based, Observational Study Using Individual-Level Data.''}
\emph{The Lancet Healthy Longevity} 1 (2): e80--88.
\url{https://doi.org/10.1016/s2666-7568(20)30016-7}.

\bibitem[\citeproctext]{ref-Brea03}
Brea, J. 2003. {``Population Dynamics in Latin America.''}
\emph{Population Bulletin} 58: 3--36.
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\bibitem[\citeproctext]{ref-chuxe1vezgalindo2016}
Chávez Galindo, Ana María, Jorge Rodríguez Vignoli, Mario Acuña, Jorge
Barquero, Daniel Macadar, José Marcos Pinto da Cunha, and Jaime Sobrino.
2016. {``Migración Interna y Cambios Metropolitanos.''} \emph{Revista
Latinoamericana de Población} 10 (18): 7--41.
\url{https://doi.org/10.31406/relap2016.v10.i1.n18.1}.

\bibitem[\citeproctext]{ref-Chen2004}
Chen, Wenhong, and Barry Wellman. 2004. {``The Global Digital Divide -
Within and Between Countries.''} \emph{IT \& Society} 1: 39--45.
\url{http://www.ifs.tuwien.ac.at/~dieter/teaching/GmA/Chen2004.pdf}.

\bibitem[\citeproctext]{ref-fielding2021}
Fielding, Tony, and Yoshitaka Ishikawa. 2021. {``COVID{-}19 and
Migration: A Research Note on the Effects of COVID{-}19 on Internal
Migration Rates and Patterns in Japan.''} \emph{Population, Space and
Place} 27 (6). \url{https://doi.org/10.1002/psp.2499}.

\bibitem[\citeproctext]{ref-firebaugh1979}
Firebaugh, Glenn. 1979. {``Structural Determinants of Urbanization in
Asia and Latin America, 1950- 1970.''} \emph{American Sociological
Review} 44 (2): 199. \url{https://doi.org/10.2307/2094505}.

\bibitem[\citeproctext]{ref-florida2021}
Florida, Richard, Andrés Rodríguez-Pose, and Michael Storper. 2021.
{``Critical Commentary: Cities in a Post-COVID World.''} \emph{Urban
Studies} 60 (8): 1509--31.
\url{https://doi.org/10.1177/00420980211018072}.

\bibitem[\citeproctext]{ref-ghosh2020}
Ghosh, Somenath, Pallabi Seth, and Harsha Tiwary. 2020. {``How Does
Covid-19 Aggravate the Multidimensional Vulnerability of Slums in India?
A Commentary.''} \emph{Social Sciences \& Humanities Open} 2 (1):
100068. \url{https://doi.org/10.1016/j.ssaho.2020.100068}.

\bibitem[\citeproctext]{ref-ginsburg2022}
Ginsburg, Carren, Mark A. Collinson, F. Xavier Gómez-Olivé, Sadson
Harawa, Chantel F. Pheiffer, and Michael J. White. 2022. {``The Impact
of COVID-19 on a Cohort of Origin Residents and Internal Migrants from
South Africa's Rural Northeast.''} \emph{SSM - Population Health} 17
(March): 101049. \url{https://doi.org/10.1016/j.ssmph.2022.101049}.

\bibitem[\citeproctext]{ref-Glaeser20}
Glaeser, Edward L. 2020. {``Cities and Pandemics Have a Long History.
City Journal.''}
\url{https://www.city-journal.org/article/cities-and-pandemics-have-a-long-history}.

\bibitem[\citeproctext]{ref-gonzuxe1lezollino2006}
González Ollino, Daniela, and Jorge Rodríguez Vignoli. 2006. {``Spatial
Redistribution and Internal Migration of the Populationin Chile over the
Past 35 Years (1965-20).''} \emph{Estudios Demográficos y Urbanos} 21
(2): 369. \url{https://doi.org/10.24201/edu.v21i2.1253}.

\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2022b}
González-Leonardo, Miguel, Antonio López-Gay, Niall Newsham, Joaquín
Recaño, and Francisco Rowe. 2022. {``Understanding Patterns of Internal
Migration During the COVID{-}19 Pandemic in Spain.''} \emph{Population,
Space and Place} 28 (6). \url{https://doi.org/10.1002/psp.2578}.

\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2023}
González-Leonardo, Miguel, Michaela Potančoková, Dilek Yildiz, and
Francisco Rowe. 2023. {``Quantifying the Impact of COVID-19 on
Immigration in Receiving High-Income Countries.''} Edited by Ricky Chee
Jiun Chia. \emph{PLOS ONE} 18 (1): e0280324.
\url{https://doi.org/10.1371/journal.pone.0280324}.

\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2022a}
González-Leonardo, Miguel, Francisco Rowe, and Alberto
Fresolone-Caparrós. 2022. {``Rural Revival? The Rise in Internal
Migration to Rural Areas During the COVID-19 Pandemic. Who Moved and
Where?''} \emph{Journal of Rural Studies} 96 (December): 332--42.
\url{https://doi.org/10.1016/j.jrurstud.2022.11.006}.

\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2022}
González-Leonardo, Miguel, and Jeroen Spijker. 2022. {``The Impact of
Covid-19 on Demographic Components in Spain, 2020{\textendash}31: A
Scenario Approach.''} \emph{Population Studies}, November, 1--17.
\url{https://doi.org/10.1080/00324728.2022.2138521}.

\bibitem[\citeproctext]{ref-graizbord2007}
Graizbord, Boris, and Beatriz Acuña. 2007. {``Movilidad Residencial En
La Ciudad de México / Residential Mobility in Mexico City.''}
\emph{Estudios Demográficos y Urbanos} 22 (2): 291.
\url{https://doi.org/10.24201/edu.v22i2.1281}.

\bibitem[\citeproctext]{ref-green2021}
Green, Mark, Frances Darlington Pollock, and Francisco Rowe. 2021.
{``New Forms of Data and New Forms of Opportunities to Monitor and
Tackle a Pandemic.''} In, 423--29. Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-70179-6_56}.

\bibitem[\citeproctext]{ref-haslag2021}
Haslag, Peter H., and Daniel Weagley. 2021. {``From L.A. To Boise: How
Migration Has Changed During the COVID-19 Pandemic.''} \emph{SSRN
Electronic Journal}. \url{https://doi.org/10.2139/ssrn.3808326}.

\bibitem[\citeproctext]{ref-huremovic2019}
Huremović, Damir. 2019. {``Brief History of Pandemics (Pandemics
Throughout History).''} In, 7--35. Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-15346-5_2}.

\bibitem[\citeproctext]{ref-internat2021}
{``International Migration 2020: Highlights.''} 2021, January.
\url{https://doi.org/10.18356/9789210052689}.

\bibitem[\citeproctext]{ref-irudayarajan2020}
Irudaya Rajan, S., P. Sivakumar, and Aditya Srinivasan. 2020. {``The
COVID-19 Pandemic and Internal Labour Migration in India: A {`}Crisis of
Mobility{'}.''} \emph{The Indian Journal of Labour Economics} 63 (4):
1021--39. \url{https://doi.org/10.1007/s41027-020-00293-8}.

\bibitem[\citeproctext]{ref-janoschka2002}
Janoschka, Michael. 2002. {``El Nuevo Modelo de La Ciudad
Latinoamericana: Fragmentación y Privatización.''} \emph{EURE
(Santiago)} 28 (85).
\url{https://doi.org/10.4067/s0250-71612002008500002}.

\bibitem[\citeproctext]{ref-kotsubo2022}
Kotsubo, Masaki, and Tomoki Nakaya. 2022. {``Trends in Internal
Migration in Japan, 2012{\textendash}2020: The Impact of the COVID{-}19
Pandemic.''} \emph{Population, Space and Place} 29 (4).
\url{https://doi.org/10.1002/psp.2634}.

\bibitem[\citeproctext]{ref-lattes95}
Lattes, A. 1995. {``Urbanización, Crecimiento Urbano y Migraciones En
América Latina. Población y Desarrollo: Tendencias y Desafíos.''}
\emph{Notas de Población} 62: 211--60.
\url{https://repositorio.cepal.org/handle/11362/38594}.

\bibitem[\citeproctext]{ref-lattes2017}
Lattes, Alfredo E., Jorge Rodríguez, and Miguel Villa. 2017.
{``Population Dynamics and Urbanization in Latin America: Concepts and
Data Limitations.''} In, 89--111. Routledge.
\url{https://doi.org/10.4324/9781315248073-5}.

\bibitem[\citeproctext]{ref-lucchini2023}
Lucchini, Lorenzo, Ollin Langle-Chimal, Lorenzo Candeago, Lucio Melito,
Alex Chunet, Aleister Montfort, Bruno Lepri, Nancy Lozano-Gracia, and
Samuel P. Fraiberger. 2023. {``Socioeconomic Disparities in Mobility
Behavior During the COVID-19 Pandemic in Developing Countries.''}
\url{https://doi.org/10.48550/ARXIV.2305.06888}.

\bibitem[\citeproctext]{ref-Maas19}
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\bibitem[\citeproctext]{ref-McKinsey21}
McKinsey \& Company. 2021. {``What Executives Are Saying about the
Future of Hybrid Work.''}
\url{https://www.mckinsey.com/capabilities/people-and-organizational-performance/our-insights/what-executives-are-saying-about-the-future-of-hybrid-work}.

\bibitem[\citeproctext]{ref-ILO2020}
Nair, Shabari, and Divya Verma. 2020. {``A Policy Framework for India's
Covid-19 Migration. International Labour Organization.''}
\url{https://www.ilo.org/newdelhi/info/public/fs/WCMS_746654/lang--en/index.htm}.

\bibitem[\citeproctext]{ref-nathan2020}
Nathan, Max, and Henry Overman. 2020. {``Will Coronavirus Cause a Big
City Exodus?''} \emph{Environment and Planning B: Urban Analytics and
City Science} 47 (9): 1537--42.
\url{https://doi.org/10.1177/2399808320971910}.

\bibitem[\citeproctext]{ref-UNpopulation19}
Nations", "United. 2019. {``World Urbanization Prospects. The 2018
Revision.''}
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\bibitem[\citeproctext]{ref-nouvellet2021}
Nouvellet, Pierre, Sangeeta Bhatia, Anne Cori, Kylie E. C. Ainslie, Marc
Baguelin, Samir Bhatt, Adhiratha Boonyasiri, et al. 2021. {``Reduction
in Mobility and COVID-19 Transmission.''} \emph{Nature Communications}
12 (1). \url{https://doi.org/10.1038/s41467-021-21358-2}.

\bibitem[\citeproctext]{ref-oecdreg}
\emph{OECD Regions and Cities at a Glance}. n.d. OECD.
\url{https://doi.org/10.1787/26173212}.

\bibitem[\citeproctext]{ref-oecdsoc}
{``OECD Social, Employment and Migration Working Papers.''} n.d.
\url{https://doi.org/10.1787/1815199x}.

\bibitem[\citeproctext]{ref-perales2022}
Perales, Francisco, and Aude Bernard. 2022. {``Continuity or Change? How
the Onset of COVID{-}19 Affected Internal Migration in Australia.''}
\emph{Population, Space and Place} 29 (2).
\url{https://doi.org/10.1002/psp.2626}.

\bibitem[\citeproctext]{ref-Puxe9rez-Campuzano13}
Pérez-Campuzano, C., E. y Santos-Cerquera. 2013. {``Tendencias Recientes
de La Migración Interna En México.''} \emph{"Papeles de Población"} 19:
53--88.
\url{https://www.scielo.org.mx/scielo.php?script=sci_arttext&pid=S1405-74252013000200003}.

\bibitem[\citeproctext]{ref-Pinilla2021}
Pinilla, Vicente, and Luis Antonio Sáez. 2021. {``La Despoblación Rural
En España: Características, Causas e Implicaciones Para Las Políticas
Públicas.''} \emph{Presupuesto y Gasto Público} 102: 75--92.
\url{https://www.ief.es/docs/destacados/publicaciones/revistas/pgp/102.pdf\#page=75}.

\bibitem[\citeproctext]{ref-PintoDaCunha02}
Pinto da Cunha, J. M. P. 2002. {``Urbanización, Redistribución Espacial
de La Población y Transformaciones Socioeconómicas En América Latina.''}
\emph{United Nations}.
\url{https://repositorio.cepal.org/bitstream/handle/11362/7168/1/S029663_es.pdf}.

\bibitem[\citeproctext]{ref-Pomeroy2021}
Pomeroy, Robin, and Ross Chiney. 2021. {``Has COVID Killed Our Cities?
World Economic Forum.''}
\url{https://www.weforum.org/agenda/2020/11/cities-podcast-new-york-dead/}.
